Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-08-04 paper-amia-summit-2018/manuscript.tex
Fixed all issues reviewed by Alex on the first draft

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by the traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose word embedding, punctuation and part-of-speech features to address it. We experimented both with traditional machine learning method, conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN). Results indicate that CRNN outperformed CRF, MLP and BRNN achieving 0.989 macro F1-score overall and 0.825 macro F1-score for detecting new segment.}

@@

\begin{enumerate}
\item Emails are unstructured text that contains informal information exchange in a non-traditional format. However, a traditional MI transcript corresponds to an MI interview
session, contains a dialogue between a counselor and a patient.
\item Discourse segments in e-Coaching emails do not necessarily correspond to sentences or collection of sentences. One sentence can be segmented into multiple MI behavior fragments. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

@@

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced word embedding or lexical, punctuation and  part-of-speech (POS) features and experimented with both traditional supervised machine learning method, linear-chain conditional random fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as multilayer perceptrons (MLP),\cite{rumelhart1986learning} bidirectional recurrent neural networks (BRNN)\cite{schuster1997bidirectional} and convolutional recurrent neural networks (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

@@

The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We formulate the segmentation task as a binary classification problem where each word or punctuation mark annotated with a label 1 or 0, to indicate whether it precedes a new segment or not. In total, we obtained 95,777 words and 7,140 punctuation marks, which include 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning method, conditional random fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as multilayer perceptrons (MLP),\cite{rumelhart1986learning} bidirectional recurrent neural networks (BRNN)\cite{schuster1997bidirectional} and convolutional recurrent neural networks (CRNN).\cite{treviso2017sentence} For MLP models, samples were created with $2l$ words or punctuations at each step in a word sequence, where a sample contains next $l$ words or punctuations and prior $l$ words or punctuations including current word or punctuation. Each sample is classified into either ``new segment'' or ``same segment'' based on whether the $l^{th}$ word or punctuation in the gold standard precedes a new segment or not. For CRF, BRNN and CRNN models, an email was taken as input sequence, such that POS tags and word embedding of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of the model at each step. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora.\cite{pennington2014glove, mikolov2013distributed} In our experiment, we utilized embeddings estimated on Google News corpus consisting of 1.6 billion words referred to pre-trained word2vec word embeddings.\footnote{https://code.google.com/p/word2vec/} When words or punctuation marks are not available in the pre-trained word vectors, we utilized word embeddings trained with our e-Coaching email corpus referred to trained word embeddings. Figure~\ref{fig:embedding-dimension} illustrates the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe and word2vec models. It was observed that best performance achieved with pre-trained 300-dimensional word2vec word vectors when three types of features are used together. Therefore, we report our results with word2vec 300-dimensional word vectors for all deep learning models used in this study. CRF utilized lexical features which represent each word in an e-Coaching email.

@@

\textbf{Multilayer Perceptrons}: MLP is a feed-forward artificial neural network which maps set of input onto a set of appropriate output.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} shows a multi layer perceptron with a single hidden layer. In an MLP networks, there are no cycles or loops where the information moves forward only, from the input nodes through the hidden nodes and to the output nodes. This study utilized MLP with a nonlinear activation function (relu) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we implement dropout on fully connected layers, which randomly hide some neurons during the training phase. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models.

@@

\section*{Results}
Our experimental results have four important directions. First, optimal sample size is identified by reporting MLP results with different sample size. Second, results are reported with respect to ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes in Table~\ref{tab:result_base}. Third, classification performance of different machine learning methods are outlined in Table~\ref{tab:result_weighted_avg} when word embeddings or lexical features are used in combination with punctuation and POS features. Fourth, the impact of individual fetures as well as their combination on the segmentation task are reported.

@@

For a given sample with $2l$ words or punctuations, the first layer input of MLP was created by the summation of first $l$ word vectors and last $l$ word vectors. Figure~\ref{fig:length-mlp} demonstrates the performance of MLP model on e-Coaching email segmentation by varying value of $l$. From the results illustrated in Figure~\ref{fig:length-mlp}, we observed that best performance is achieved when $l = 2$. Therefore, all MLP results are reported with $l = 2$ in the remaining part of this section.

@@

Table~\ref{tab:result_weighted_avg} summarizes the results of all models for segmentation of e-Coaching emails when word embeddings or lexical features are used in combination with punctuation and POS features. Similar to results in Tables~\ref{tab:result_base}, CRNN demonstrates the highest performance among all methods achieving 0.864 recall with 0.825 F1-Measure for ``new segment'' class and 0.990 precision with 0.989 recall and F1-Measure for overall. BRNN and CRF show the lowest and second highest performance, respectively, for email segmentation among all methods. We observed that classification performance significantly improved for ``new segment'' class when word embeddings or lexical features are used in combination with punctuation and POS features. Precision increases by 3.96\%, -2.27\%, 12.71\% and 1.81\%; recall increases by 11.72\%, 19.73\%, 20.59\% and 8.41\%; and F1-measure increases by 8.05\%, 9.51\%, 16.22\% and 5.1\% for CRF, MLP, BRNN and CRNN methods, respectively, in new segment detection when all features are utilized together. Similarly, precision increases by 0.51\%, 0.41\%, 0.82\% and 0.41\%; recall increases by 0.41\%, 0.41\%, 0.72\% and 0.3\%; and F1-measure increases by 0.41\%, 0.41\%, 0.82\% and 0.3\% for CRF, MLP, BRNN and CRNN methods, respectively, in weighted average over ``new segment'' and ``same segment'' classes when word embeddings or lexical features are used in combination with punctuation and POS features.

Table~\ref{tab:result_roc} shows the impact of individual features as well as their combination on the segmentation task. Although CRF provides the second highest classification results, it shows best AUC values, achieving 0.966 AUC when only lexical features are used and 0.994 AUC when a combination of lexical, punctuation and POS features are used. Influence of the punctuation and POS features is also consistent in AUC values, which increase by 2.9\%, 4.31\%, 6.63\% and 2.18\% for CRF, MLP, BRNN and CRNN methods, respectively. Individually, POS and punctuation features improve the performance of all classifiers except CRNN when word embeddings are used with POS features. CRF and MLP achieve their highest AUCs when all features are used together. On the other hand, BRNN and CRNN demonstrate the highest AUCs when word embeddings are combined with punctuation features. \\

\begin{table}[ht]
\centering
\caption{\textbf{AUC values of all classifiers when only word embeddings or lexical features and combination of word embeddings or lexical, punctuation and POS features are used. Highest value for each experiment across all models is highlighted in boldface.}}
\label{tab:result_roc}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\multirow{2}{*}{\textbf{Features}} & \multicolumn{4}{|c|}{\textbf{AUC}} \\\cline{2-5}
 & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings only & \textbf{0.966} & 0.951 & 0.920 & 0.965 \\ \hline
 word embeddings + POS & \textbf{0.973} & 0.953 & 0.920 & 0.962 \\ \hline
 word embeddings + punctuation & \textbf{0.993} & 0.989 & 0.982 & 0.988 \\ \hline
 all features & \textbf{0.994} & 0.992 & 0.981 & 0.986 \\ \hline
  \end{tabular}
\end{table}

@@

>>>>

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by the traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose word embeddings, punctuation and part-of-speech features to address it. We experimented both with traditional machine learning method, conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN). Results indicate that CRNN outperformed CRF, MLP and BRNN achieving 0.989 macro F1-score overall and 0.825 macro F1-score for detecting new segment.}

@@

\begin{enumerate}
\item Emails are unstructured text that contains informal information exchange in a non-traditional format. For example, an e-Coach usually responds to several queries of a patient in one email. However, a provider utterance of MI transcript denotes a particular response for a query of a patient during a traditional MI interview session, which involve a dialogue between a provider and a patient. 
\item Discourse segments in e-Coaching emails do not necessarily correspond to sentences or collection of sentences. One sentence can be segmented into multiple MI behavior fragments. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

@@

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced word embeddings or lexical, punctuation and  part-of-speech (POS) features and experimented with both traditional supervised machine learning method, linear-chain conditional random fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as multilayer perceptrons (MLP),\cite{rumelhart1986learning} bidirectional recurrent neural networks (BRNN)\cite{schuster1997bidirectional} and convolutional recurrent neural networks (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

@@

The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We formulate the segmentation task as a binary classification problem where each word or punctuation mark annotated with a label 1 or 0, to indicate whether it precedes a new segment or not. In total, we obtained 95,777 words and 7,140 punctuation marks, which include 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning method, conditional random fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as multilayer perceptrons (MLP),\cite{rumelhart1986learning} bidirectional recurrent neural networks (bidirectional RNN or simply BRNN)\cite{schuster1997bidirectional} and convolutional recurrent neural networks (CRNN).\cite{treviso2017sentence} For MLP models, samples were created with $2l$ words or punctuations at each step in a word sequence, where a sample contains next $l$ words or punctuations and prior $l$ words or punctuations including current word or punctuation. Each sample is classified into either ``new segment'' or ``same segment'' based on whether the $l^{th}$ word or punctuation in the gold standard precedes a new segment or not. For CRF, BRNN and CRNN models, an email was taken as input sequence, such that POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of the model at each step. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora.\cite{pennington2014glove, mikolov2013distributed} In our experiment, we utilized embeddings estimated on Google News corpus consisting of 1.6 billion words referred to pre-trained word2vec word embeddings.\footnote{https://code.google.com/p/word2vec/} When words or punctuation marks are not found in the pre-trained word vectors, we utilized word embeddings trained with our e-Coaching email corpus referred to trained word embeddings. CRF utilized lexical features which represent each word in an e-Coaching email.

@@

\textbf{Multilayer Perceptrons}: MLP is a feed-forward artificial neural network which maps set of input onto a set of appropriate output.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} shows a multilayer perceptron with a single hidden layer. In MLP networks, there is no cycles or loops and information moves forward only, from the input nodes through the hidden nodes and to the output nodes. This study utilized MLP with a nonlinear activation function (relu) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we implement dropout on fully connected layers, which randomly hide some neurons during the training phase. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models.

@@

\section*{Results}
Our experimental results have four important directions. First, the optimal dimension of word embeddings and sample size of MLP models are reported. Second, results are reported with respect to ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes in Table~\ref{tab:result_base}. Third, classification performance of different machine learning methods are outlined in Table~\ref{tab:result_weighted_avg} when word embeddings or lexical features are used in combination with punctuation and POS features. Fourth, the impact of the individual, as well as their combination on the segmentation task, are reported.

Figure~\ref{fig:embedding-dimension} illustrates the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe and word2vec models. It was observed that best performance achieved with pre-trained 300-dimensional word2vec word vectors when three types of features are used together. Therefore, we report our results with word2vec 300-dimensional word vectors for all deep learning models used in this study. For MLP models, the first layer input was prepared by the summation of first $l$ word vectors and last $l$ word vectors when a sample contains $2l$ words or punctuations. Figure~\ref{fig:length-mlp} demonstrates the performance of MLP model on e-Coaching email segmentation by varying the value of $l$. It was observed that the best performance of MLP is achieved when $l = 2$. Therefore, results of all MLP are reported with $l = 2$ in the remaining part of this section.

@@

@@

Table~\ref{tab:result_weighted_avg} summarizes the results of all models for segmentation of e-Coaching emails when word embeddings or lexical features are used in combination with punctuation and POS features. Similar to results in Tables~\ref{tab:result_base}, CRNN demonstrates the highest performance among all methods achieving 0.864 recall with 0.825 F1-Measure for ``new segment'' class and 0.990 precision with 0.989 recall and F1-Measure for overall. BRNN and CRF show the lowest and second highest performance, respectively, for email segmentation among all methods. We observed that classification performance significantly improved for ``new segment'' class when word embeddings or lexical features are used in combination with punctuation and POS features. Precision increases by 3.96\%, -2.27\%, 12.71\% and 1.81\%; recall increases by 11.72\%, 19.73\%, 20.59\% and 8.41\%; and F1-measure increases by 8.05\%, 9.51\%, 16.22\% and 5.1\% for CRF, MLP, BRNN and CRNN methods, respectively, in new segment detection when all features are utilized together. Similarly, precision increases by 0.51\%, 0.41\%, 0.82\% and 0.41\%; recall increases by 0.41\%, 0.41\%, 0.72\% and 0.3\%; and F1-measure increases by 0.41\%, 0.41\%, 0.82\% and 0.3\% for CRF, MLP, BRNN and CRNN methods, respectively, in weighted average over ``new segment'' and ``same segment'' classes when word embeddings or lexical features are used in combination with punctuation and POS features.\\

\begin{table}[ht]
\centering
\caption{\textbf{AUC values of all classifiers demonstrating the impact of word embeddings, punctuation and POS features on email segmentation. Highest AUC value for each feature set across all models is highlighted in boldface.}}
\label{tab:result_roc}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\multirow{2}{*}{\textbf{Features}} & \multicolumn{4}{|c|}{\textbf{AUC}} \\\cline{2-5}
 & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings only & \textbf{0.966} & 0.951 & 0.920 & 0.965 \\ \hline
 word embeddings + POS & \textbf{0.973} & 0.953 & 0.920 & 0.962 \\ \hline
 word embeddings + punctuation & \textbf{0.993} & 0.989 & 0.982 & 0.988 \\ \hline
 all features & \textbf{0.994} & 0.992 & 0.981 & 0.986 \\ \hline
  \end{tabular}
\end{table}

@@

Table~\ref{tab:result_roc} shows the impact of individual features as well as their combination on the segmentation task. Although CRF provides the second highest classification results, it shows best AUC values, achieving 0.966 AUC when only lexical features are used and 0.994 AUC when a combination of lexical, punctuation and POS features are used. Influence of the punctuation and POS features is also consistent in AUC values, which increase by 2.9\%, 4.31\%, 6.63\% and 2.18\% for CRF, MLP, BRNN and CRNN methods, respectively. Individually, POS and punctuation features improve the performance of all classifiers except CRNN when word embeddings are used with POS features. CRF and MLP achieve their highest AUCs when all features are used together. On the other hand, BRNN and CRNN demonstrate the highest AUCs when word embeddings are combined with punctuation features.
